NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2019-10-29 Manuscript/manuscriptPGPE.tex
Add CHARMM PCvsPG results without ions
Update texty

\subsection{PC headgroup interactions with PE and PG}
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPEPG.eps}
  \caption{\label{HGorderparametersPCvsPEPG}
    Modulation of POPC headgroup order parameters with increasing amount of POPE (left) and POPG (right) in bilayer
    from experiments \cite{scherer87,macdonald87} and simulations with different force fields.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
  \todo{Data for CHARMM without ions to be added once this issue is solved: https://github.com/NMRLipids/MATCH/issues/83}
  %\todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.}
\end{figure*}

In experiments, the PC headgroup order parameters increase
with the addition of negatively charged PG or PS lipids, but
are not affected by the addition of zwitterionic PE and SM lipids
or cholesterol (Fig.~\ref{HGorderparametersPCvsPEPG}).
This can be explained by the electrometer concept, which suggests that the 
headgroup dipole tilts more parallel to the membrane plane upon addition of
negative charge to the membrane~\cite{seelig87, scherer87,antila18}.
The different responses of PC headgroup order parameters to PE and PG are roughly
reproduced in CHARMM36 simulations, although the changes are slightly
overestimated upon addition of PE and underestimated upon addition of PG (Fig.~\ref{HGorderparametersPCvsPEPG}).
Interestingly, the response of POPC headgroup order parameters to POPE is
close to experiments also in Berger-OPLS simulations, even though the response
of Berger force field to cholesterol was significantly overestimated in our previous work \cite{botan15}.
In all force fields except Slipids, the $\alpha$-carbon order parameters of different
hydrogens are responsing differently when mixed with PE or PG lipids \todo{Maybe we should figure out what is the reason for this}.
\todo{Maybe we should analyze the P-N vector angle from different simulations}.

For $\beta$-carbon order parameter in PG headgroup, experiments report
mild increase \cite{macdonald87} or no change~\cite{borle85} upon addition 
of PC lipids (Fig. \ref{HGorderparametersPGvsPCchange}). 
Simulations with all the tested force fields give only very small changes also for
the $\alpha$-carbon order parameter (Figs. \ref{HGorderparametersPGvsPC} and \ref{HGorderparametersPGvsPCchange}). 
Therefore, the simulations are generally in line with experimetns, suggesting that the
interactions with PC do not essentially effect the PG headgroup structure.
This is in contrast with previous results for PS headgroup \cite{antila18}, where
all the force fieds significantly overestimated the structural response of PS headgroup
to the interactions with PC lipids.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPGvsPCchange.eps}
  \caption{\label{HGorderparametersPGvsPCchange}
    Modulation of PG lipid headgroup order parameters with the increasing amount of PC in lipid bilayer
    from experiments \cite{borle85,macdonald87} and simulations with different force fields.
  }
\end{figure}

@@

\subsection{Sodium binding to PE and PG lipid bilayers}
In our previous study about PS lipids \cite{antila18}, the underestimated increase of PC
headgroup order parameters upon addition of negaticely charged lipids
was related
to the overestimated counterion binding affinity, which would overcompensates the effect of negative charge.
Similar correlation is observed here in %other than GROMOS-CKP simulation:
Lipid 17 simulations, which exhibits stronger counter-ion binding affinity to pure POPG bilayer (Fig. \ref{CIdensPG})
and larger decrease in POPC $\beta$ carbon order parameter upon addition of POPG than other simulations.
Together with the lower area per molecule (59.5 {\AA}$^2$) than in experiments (66.1 {\AA}$^2$),
the results suggest that the counterions bind too strongly and shield the electrostatic repulsion between PG headgroups
in bilayers simulated with Lipid17 parameters.
In GROMOS-CKP simulations, however, the $\beta$-carbon order parameter of POPC
increases upon addition of POPG more than in CHARMM36 and Slipids simulations even though the binding affinities in these
simulations are similar. On the other hand, the responses of different hydrogens attached in $\alpha$-carbons
are different in all simulations except in Slipids, most sever in Lipid17 and GROMOS-CKP where other in increases
and other decreases \todo{Can we figure out the reason for this?}.
In conclusion, the counter-ion binding seems to be overestimated in Lipid17 with the strongest affinity among the
tested models, while with this data we cannot fully evaluate binding affinity in the other models.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CIdensPG.eps}
  \caption{\label{CIdensPG}
    Counterion densities and area per lipids from simulations with PG lipids.
    Experimental area for POPG at 303 K is 66.1 {\AA}$^2$ and 67 {\AA}$^2$ for DPPC at 323 K \cite{pan12b}.
  }
\end{figure}

%Sodium binding to PC lipid bilayers is significantly overestimated by most simulations models \cite{catte16}.
%Also, sodium binding to PS lipids seems to be overestimated, although the
%presence of counterions complicate the comparison for negatively charged lipids \cite{antila19}.
Similarly to POPG, the Lipid17 exhibits stronger sodium binding affinity
to POPE bilayers than Slipids, CHARMM36 and GROMOS-CKP (Fig. \ref{NAdensPE}).
In Lipid17 the sodium binding affinity to POPE is approximately similar as to POPC,
while in other models exhibit weaker binding to POPE (Fig. \ref{NAdensPC}).
The difference is small in Slipids and CHARMM36, while GROMOS-CKP predicts
subtantially stronger binding to POPC.
Experimental data for sodium binding to PE lipids is not available, while
simulations in agreement with electrometer data from NMR suggests that
sodium binding to POPC is weaker than in any of the simulations
here (Fig. \ref{NAdensPC})~\cite{catte16,melcr18}. Assuming that the binding
to POPE would be similar than to POPC, the sodium binding affinity to POPE
is potentially realistic in CHARMM36, Slipids, and GROMOS-CKP simulations here,
but substantially overestimated in Lipid17 simulation.
%On the other hand, the sodium binding affinity is stronger to POPE in Lipid17 simulations than to POPC in simulations with
%previous Amber lipid version, Lipid14.
\todo{The potential differences in the used ion models should be checked}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/NAdensPE.eps}
  \caption{\label{NAdensPE}
    Sodium (solid line) and choride ion density profiles along membrane normal
    from different simulations with PE lipids.
  }
\end{figure}

@@

\section{Conclusions}

>>>>

\clearpage
\subsection{PC headgroup interactions with PE and PG}
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPEPG.eps}
  \caption{\label{HGorderparametersPCvsPEPG}
    Modulation of POPC headgroup order parameters with increasing amount of POPE (left) and POPG (right) in bilayer
    from experiments \cite{scherer87,macdonald87} and simulations with different force fields.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
  %\todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.}
\end{figure*}

In experiments, the PC headgroup order parameters increase
with the addition of negatively charged PG or PS lipids, but
are not affected by the addition of zwitterionic PE and SM lipids
or cholesterol (Fig.~\ref{HGorderparametersPCvsPEPG}).
This can be explained by the electrometer concept, which suggests that the 
headgroup dipole tilts more parallel to the membrane plane upon addition of
negative charge to the membrane~\cite{seelig87, scherer87,antila18}.
The response of PC headgroup order parameters to PE by the tested CHARMM36 and
Berger-OPLS force fields, although CHARMM36 slightly overestimates the changes (Fig.~\ref{HGorderparametersPCvsPEPG}).
The good performance of Berger-OPLS simulations is notable because the response of headgroup order parameters
to cholesterol was significantly overestimated by the Berger/Höltje force field in our previous work \cite{botan15}.
None of the force fields fully reproduces the PC headgroup order parameter response to the increasing amount of
PG, which may be related to the counterion binding affinity (see also the next section) \cite{antila19}.
In all force fields except Slipids, the order parameters of different
hydrogens attached to the $\alpha$-carbon are responsing differently when mixed with PE or PG lipids
\todo{Maybe we should figure out what is the reason for this?}.\\
\todo{Maybe we should analyze the P-N vector angle from different simulations?}.

For $\beta$-carbon order parameter in PG headgroup, experiments report
mild increase \cite{macdonald87} or no change~\cite{borle85} upon addition 
of PC lipids (Fig. \ref{HGorderparametersPGvsPCchange}). 
Simulations with all the tested force fields give only very small changes also for
the $\alpha$-carbon order parameter (Figs. \ref{HGorderparametersPGvsPC} and \ref{HGorderparametersPGvsPCchange}). 
Therefore, the simulations are generally in line with experiments, suggesting that the
interactions with PC do not essentially effect the PG headgroup structure.
This suggests that the interactions between PG and PC headgroups are captured better in
simulations than for PS headgroup, where
all the force fieds significantly overestimated the structural response of PS headgroup
to the interactions with PC lipids \cite{antila18}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPGvsPCchange.eps}
  \caption{\label{HGorderparametersPGvsPCchange}
    Modulation of PG lipid headgroup order parameters with the increasing amount of PC in lipid bilayer
    from experiments \cite{borle85,macdonald87} and simulations with different force fields.
  }
\end{figure}

@@

\clearpage
\subsection{Sodium binding to PE and PG lipid bilayers}
Sodium binding affinity to PE lipids has not been measured experimentally,
but large differences to PC would be surprising. 
The sodium binding affinity to POPE depends on the used force field (Fig. \ref{NAdensPE}),
but lesser extend than reported previously for PC~\cite{catte16}.
%Similarly to POPG,
Because some simulation and ion parameters are not identical with the previous work~\cite{catte16}
\todo{This should be finished once we have the simulation details},
we compare POPE results to the POPC simulations ran with identical parameters (Fig. \ref{NAdensPC}).
In Lipid17 with the strongest sodium binding affinity to POPE,
the binding affinity is approximately similar to POPC.
Slipids and CHARMM36 exhibit slightly, and
GROMOS-CKP subtantially weaker binding to POPE than to POPC.
%The Lipid17 exhibits stronger sodium binding affinity
%to POPE bilayers than Slipids, CHARMM36 and GROMOS-CKP (Fig. \ref{NAdensPE}).
%Experimental data for sodium binding to PE lipids is not available, while
%simulations in agreement with electrometer data from NMR suggests that
%sodium binding to POPC is weaker than in any of the simulations
%here (Fig. \ref{NAdensPC})~\cite{catte16,melcr18}.
Assuming that the binding
to POPE would be similar than to POPC, the sodium binding affinity to POPE
is potentially realistic in CHARMM36, Slipids, and GROMOS-CKP simulations here,
but substantially overestimated in Lipid17 simulation.
%On the other hand, the sodium binding affinity is stronger to POPE in Lipid17 simulations than to POPC in simulations with
%previous Amber lipid version, Lipid14.
%\todo{The potential differences in the used ion models should be checked}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/NAdensPE.eps}
  \caption{\label{NAdensPE}
    Sodium (solid line) and choride ion density profiles along membrane normal
    from different simulations with PE lipids.
  }
\end{figure}

Simulations with PG lipids give similar dependence on force field as observed in POPE simulations:
Lipid 17 simulations exhibits stronger counter-ion binding affinity to pure POPG bilayer
than CHARMM36, Slipids, and GROMOS-CKP simulations, which are roughly similar (Fig. \ref{CIdensPG}).
Lipid17 also exhibits less increase in POPC headgroup
%$\beta$ carbon
order parameters upon addition of POPG than other simulations (Fig. \ref{HGorderparametersPCvsPEPG}),
and lower area per molecule (59.5 {\AA}$^2$) than in experiments (66.1 {\AA}$^2$).
In our previous study about PS lipids \cite{antila18}, such behaviour was related to 
%the results suggest that
the overestimated counterions binding and shielding the electrostatic repulsion between PG headgroups
in bilayers.% simulated with Lipid17 parameters.
%the underestimated increase of PC
%headgroup order parameters upon addition of negaticely charged lipids
%was related
%to the overestimated counterion binding affinity, which would overcompensates the effect of negative charge.
%Similar correlation is observed here in %other than GROMOS-CKP simulation:
%%Lipid 17 simulations, which exhibits stronger counter-ion binding affinity to pure POPG bilayer (Fig. \ref{CIdensPG})
%
Even though the area per lipid in CHARMM36, Slipids, and GROMOS-CKP simulations is in good agreement
with experiments (Fig. \ref{CIdensPG}), the experimental increase in 
%In GROMOS-CKP simulations, however, the $\beta$-carbon order parameter of
POPC headgroup order parameters upon addition of POPG are not fully reproduced (Fig. \ref{HGorderparametersPCvsPEPG}).
%more than in CHARMM36 and Slipids simulations even though the binding affinities in these
%simulations are similar. On the other hand, the responses of different hydrogens attached in $\alpha$-carbons
%are different in all simulations except in Slipids, most sever in Lipid17 and GROMOS-CKP where other in increases
%and other decreases \todo{Can we figure out the reason for this?}.
Therefore we conclude that the counter-ion binding affinity is overestimated in Lipid17 simulations,
% with the strongest affinity among the tested models,
while the other simulations are more realistic, but slight overbinding cannot be excluded.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CIdensPG.eps}
  \caption{\label{CIdensPG}
    Counterion densities and area per lipids from simulations with PG lipids.
    Experimental area for POPG at 303 K is 66.1 {\AA}$^2$ and 67 {\AA}$^2$ for DPPC at 323 K \cite{pan12b}.
  }
\end{figure}

@@

\clearpage
\section{Conclusions}
